\Title

### 2021-03-02 latex/paper/main.tex
First full draft of the VNHC section of the paper

In light of Lemma \ref{lemma:B-orthogonal}, there is no loss of generality in
assuming that the input matrix is left semi-orthogonal.
Let now \(\mathbf{B} = [B^\perp; B^\tpose]\).
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal.

@@

The authors of
\cite{nhvc_dynamic_walking,hybrid_zero_dynamics_bipedal_nhvcs,nhvc_incline_walking}
observed that only relations which use the unactuated conjugate of momentum
have vector relative degree \(\{2,\ldots,2\}\).
However, this is not enough to guarantee the

We now provide a computational characterization of regularity
we define the following matrices as shorthand:
\begin{align}
    \mathcal{M}(q,p) &:= (\Id{n-k} \otimes p\tpose)\nabla_{q_u}\Minv(q) 
        ,\\
    \mathcal{D}(q,p_u) &:= dh_q(q,p_u) \Minv(q) 
    .
\end{align}

\begin{thm}\label{thm:vnhc-regularity}
    A relation \(h(q,p) = 0\) for system ~\eqref{eqn:simple-hamiltonian}
    is a regular VNHC of order \(k\) if and only if \(dh_{p_a} = 0\) 
    and
    \[
        \rank{\left(\mathcal{D}(q,p_u) - dh_{p_u}\mathcal{M}(q,p)\right)\simpleB} = k
         ,
    \]
    everywhere on the constraint manifold \(\Gamma\).
\end{thm}
\begin{proof}
    Let \(e = h(q,p) \in \R^k\).
    If \(dh_{p_a} \neq \Zmat{k\times k}\) for some \((q,p) \in \Gamma\), 
    then \(\tau\) appears in \(\dot{e}\) and the VNHC is not of relative degree
    \(\{2,\ldots,2\}\). Suppose now that \(dh_{p_a} = \Zmat{k\times k}\).
    Then 
    \[
        \dot{e} = \mathcal{D}(q,p_u)p - 
        dh_{p_u}\left(\frac{1}{2}\mathcal{M}(q,p)p + \nabla_{q_u}V(q)\right)
        .
    \]
    Taking one further derivative provides
    \begin{align*}
        \ddot{e} &= (\star) - 
        dh_{p_u}\left(\frac{1}{2}\diff{}{t}\left(\mathcal{M}(q,p)p\right)\right) 
        \\
                 &+\mathcal{D}(q,p_u)\simpleB \tau
    \end{align*}
    where \((\star)\) is a continuous function of \(q\) and \(p\).
    One can show that
    \begin{align*}
        dh_{p_u}\left(\frac{1}{2}\diff{}{t}\left(\mathcal{M}(q,p)p\right)\right)
        &= (\star) +\\
        &dh_{p_u} \mathcal{M}(q,p)\simpleB \tau
        .
    \end{align*}
    Hence,
    \[
       \ddot{e} = (\star) +
       \left(\mathcal{D}(q,p_u) - dh_{p_u}\mathcal{M}(q,p)\right) \simpleB \tau
        ,
    \]
    which we write as \( \ddot{e} = E(q,p) + H(q,p)\tau\) for appropriate \(E\)
    and \(H\).
    From the definition of regularity, the VNHC \(h\) is regular 
    when \(e\) is of relative degree \(\{2,\ldots,2\}\), which is true 
    if and only if the matrix premultiplying \(\tau\) is nonsingular, and hence
    that \(H\) is invertible. This proves the theorem.
\end{proof}

The constrained dynamics for a general VNHC were derived by 
Horn et. al \cite{nhvc_dynamic_walking}.
In our case, we will be enforcing regular constraints where the actuators of the
acrobot track a function of the unactuated variables. 
Intuitively, the dynamics should be parameterized by \((q_u, p_u)\).
Unfortunately, \(\dot{q}_u\) depends on \(p_a\), and for general systems one
cannot solve explicitly for \(p_a\) in terms of \((q_u,p_u)\). 
This is because the \(\dot{p}\) dynamics contains the coupling term 
\((\Id{n} \otimes p\tpose)\nabla_{q_u}M(q)p\).

We now introduce an assumption so we can solve explicitly for the constrained
dynamics.

\begin{assm}\label{assm:inertially-actuated}
    The Hamiltonian system has an inertia matrix that does not depend on the
    unactuated coordinates, so that 
    \(\nabla_{q_u}M(q) = \Zmat{n(n-k) \times n}\).
\end{assm}

\begin{thm}\label{thm:zero-dynamics}
    Let \(\mathcal{H}\) be a mechanical system in simply actuated
    coordinates satisfying Assumption \ref{assm:inertially-actuated}. 
    Let \(h(q,p_u) = 0\) be a regular VNHC of order \(k\) with constraint
    manifold \(\Gamma\). Suppose that on \(\Gamma\) one can solve for
    \(q_a\) as a function \(q_a = f(q_u,p_u)\).
    Then the constrained dynamics are given by
    \begin{equation}\label{eqn:qpu-dynamics}
        \left.\begin{aligned}
                \dot{q}_u &= \begin{bmatrix}
                    \Id{(n-k)} & \Zmat{(n-k) \times k}
                \end{bmatrix}\Minv(q)p \\
            \dot{p}_u &= -\nabla_{q_u}V(q) \\
            \end{aligned}{}\right|_{\begin{array}{c}
                q_a = f(q_u,p_u) \\ 
                p_a = g(q_u,p_u) \\
            \end{array}}
            ,
    \end{equation}
    where
    \begin{equation}\label{eqn:g-qpu}
    \begin{aligned}
        &g(q_u,p_u) := \\
        &\left(\mathcal{D}(q,p_u) \simpleB \right)\inv 
               \\
        &\left.\left(dh_{p_u} \nabla_{q_u}V(q) - \mathcal{D}(q,p_u)
        \begin{bmatrix}
            I_{n-k} \\
            \Zmat{k \times (n-k)} \\
    \end{bmatrix} p_u\right)\right|_{qa = f(q_u,p_u)}
        .
    \end{aligned}
    \end{equation}
\end{thm}
\begin{proof}
    This is a special case of Theorem 1 in \cite{nhvc_dynamic_walking} using
    \(\theta_1 = q_u\) and \(\theta_2 = p_u\).
    The proof is ommitted.
\end{proof}

>>>>

In light of Lemma \ref{lemma:B-orthogonal}, there is no loss of generality in
assuming that the input matrix is left semi-orthogonal.
Let now \(\mathbf{B} = [B^\perp; B\tpose]\).
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal.

@@

The authors of
\cite{nhvc_dynamic_walking,hybrid_zero_dynamics_bipedal_nhvcs,nhvc_incline_walking}
observed that relations which use only the unactuated conjugate of momentum
often have vector relative degree \(\{2,\ldots,2\}\).
Indeed, we shall now provide a characterization of regularity which shows that
regular constraints cannot use the actuated momentum at all.

To ease notation in the rest of this section, we use the following shorthand:
\begin{align}
    \mathcal{M}(q,p) &:= (\Id{n-k} \otimes p\tpose)\nabla_{q_u}\Minv(q) 
        ,\\
    \mathcal{D}(q,p_u) &:= dh_q(q,p_u) \Minv(q) 
    .
\end{align}

\begin{thm}\label{thm:vnhc-regularity}
    A relation \(h(q,p) = 0\) for system ~\eqref{eqn:simple-hamiltonian}
    is a regular VNHC of order \(k\) if and only if \(dh_{p_a} = 0\) 
    and
    \[
        \rank{\left(\mathcal{D}(q,p_u) - dh_{p_u}\mathcal{M}(q,p)\right)\simpleB} = k
         ,
    \]
    everywhere on the constraint manifold \(\Gamma\).
\end{thm}
\begin{proof}
    Let \(e = h(q,p) \in \R^k\).
    If \(dh_{p_a} \neq \Zmat{k\times k}\) for some \((q,p) \in \Gamma\), 
    then \(\tau\) appears in \(\dot{e}\) and the VNHC is not of relative degree
    \(\{2,\ldots,2\}\). Suppose now that \(dh_{p_a} = \Zmat{k\times k}\).
    Then 
    \[
        \dot{e} = \mathcal{D}(q,p_u)p - 
        dh_{p_u}\left(\frac{1}{2}\mathcal{M}(q,p)p + \nabla_{q_u}V(q)\right)
        .
    \]
    Taking one further derivative provides
    \begin{align*}
        \ddot{e} 
            &= (\star) - 
            dh_{p_u}\left(\frac{1}{2}\diff{}{t}\left(\mathcal{M}(q,p)p\right)\right) 
                \\
            &+ \mathcal{D}(q,p_u)\simpleB \tau
    \end{align*}
    where \((\star)\) is a continuous function of \(q\) and \(p\).
    One can show that
    \begin{align*}
        dh_{p_u}\left(\frac{1}{2}\diff{}{t}\left(\mathcal{M}(q,p)p\right)\right)
        &= (\star) +\\
        &dh_{p_u} \mathcal{M}(q,p)\simpleB \tau
        .
    \end{align*}
    Hence,
    \[
       \ddot{e} = (\star) +
       \left(\mathcal{D}(q,p_u) - dh_{p_u}\mathcal{M}(q,p)\right) \simpleB \tau
        ,
    \]
    which we write as \( \ddot{e} = E(q,p) + H(q,p)\tau\) for appropriate \(E\)
    and \(H\).
    From the definition of regularity, the VNHC \(h\) is regular 
    when \(e\) is of relative degree \(\{2,\ldots,2\}\), which is true 
    if and only if the matrix premultiplying \(\tau\) is nonsingular, and hence
    that \(H\) is invertible. This proves the theorem.
\end{proof}

Under additional mild conditions (see \cite{vhcs_for_el_systems}), a regular VNHC of
order \(k\) can be stabilized by the output-linearizing phase-feedback
controller
\begin{equation}
    \tau(q,p) = -H\inv(q,p)\left(E(q,p) + k_p e + k_d \dot{e}\right)
    ,
\end{equation}
where \(k_p, k_d > 0\) are control parameters which can be tuned on the
resulting error dynamices \(\ddot{e} = -k_p e - k_d \dot{e}\).

In Section \ref{sec:acrobot} we will enforce a regular constraint on the
acrobot of the form \(h(q,p) = q_a - f(q_u,p_u)\), where the actuators track a
function of the unactuated variables.
Intuitively then, the constrained dynamics should be parameterized by \((q_u, p_u)\).
Unfortunately, \(\dot{q}_u\) depends on \(p_a\), and for general systems one
cannot solve explicitly for \(p_a\) in terms of \((q_u,p_u)\) because
the \(\dot{p}\) dynamics contains the coupling term 
\((\Id{n} \otimes p\tpose)\nabla_{q_u}M(q)p\).

We now introduce an assumption which allows us to solve for \(p_a\) as a
function of \((q_u,p_u)\), which in turn allows us to find the constrained
dynamics.

\begin{assm}\label{assm:inertially-actuated}
The inertia matrix does not depend on the unactuated coordinates, so that 
\(\nabla_{q_u}M(q) = \Zmat{n(n-k) \times n}\).
\end{assm}

\begin{thm}\label{thm:zero-dynamics}
    Let \(\mathcal{H}\) be a mechanical system in simply actuated
    coordinates satisfying Assumption \ref{assm:inertially-actuated}. 
    Let \(h(q,p_u) = 0\) be a regular VNHC of order \(k\) with constraint
    manifold \(\Gamma\). Suppose that on \(\Gamma\) one can solve for
    \(q_a\) as a function \(q_a = f(q_u,p_u)\).
    Then the constrained dynamics are given by
    \begin{equation}\label{eqn:qpu-dynamics}
        \left.\begin{aligned}
                \dot{q}_u &= \begin{bmatrix}
                    \Id{(n-k)} & \Zmat{(n-k) \times k}
                \end{bmatrix}\Minv(q)p \\
            \dot{p}_u &= -\nabla_{q_u}V(q) \\
            \end{aligned}{}\right|_{\begin{array}{c}
                q_a = f(q_u,p_u) \\ 
                p_a = g(q_u,p_u) \\
            \end{array}}
            ,
    \end{equation}
    where
    \begin{equation}\label{eqn:g-qpu}
    \begin{aligned}
        &g(q_u,p_u) := \\
        &\left(\mathcal{D}(q,p_u) \simpleB \right)\inv 
               \\
        &\left.\left(dh_{p_u} \nabla_{q_u}V(q) - \mathcal{D}(q,p_u)
        \begin{bmatrix}
            I_{n-k} \\
            \Zmat{k \times (n-k)} \\
    \end{bmatrix} p_u\right)\right|_{qa = f(q_u,p_u)}
        .
    \end{aligned}
    \end{equation}
\end{thm}
\begin{proof}
    Setting \(e = h(q,p_u)\) and using Assumption
    \ref{assm:inertially-actuated}, we find that
    \(\dot{e} = \mathcal{D}(q,p_u)p - dh_{p_u}\nabla_{q_u}V(q)\).
    Notice that
    \(\mathcal{D}(q,p_u)p = \mathcal{D}(q,p_u)[\Zmat{(n-k)\times k}; \Id{k}]p_a
    + \mathcal{D}(q,p_u)[\Id{n-k};\Zmat{k \times (n-k)}] p_u\).
    Since \(h(q,p_u)\) is regular, \(\mathcal{D}(q,p_u)\) is invertible.
    Taking \(e = \dot{e} = 0\), solving for \(p_a\), and setting 
    \(q_a = f(q_u,p_u)\) completes the proof.
\end{proof}

\textbf{Comparison with existing literature}: Horn et.al. provide constrained
dynamics for VNHCs in \cite{nhvc_incline_walking}.
In fact, our constrained dynamics \eqref{eqn:qpu-dynamics} coincide with their
system (17) when one chooses the special case \(\theta_1 = q_u\) and 
\(\theta_2 = p_u\).
Other than this distinction, our assumptions are identical to theirs.
In particular, their assumption \textbf{H2} is what we call
regularity.
The only real distinction is that our constrained dynamics
are explicit functions of the Hamiltonian phase coordinates
\((q_u,p_u)\).
This representation will be beneficial when we apply the theory of VNHCs to the
acrobot.
